Climate change projected to reduce European life expectancy

### 2018-04-05 MS-cclifeexpec.Rmd
blinded.

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "Climate change projected to reduce European life expectancy"
thanks: "All data and code that supports these conclusions are available as supplementary materials."
author:
- name: Mathew E. Hauer^1^*
  affiliation: University of Georgia
- name: Alexis R. Santos^2^
  affiliation: Pennsylvania State University

abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change. However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups. No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy. Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality. In this paper, we convert excess mortality estimates into potential reductions of life expectancy at birth and report the cost of climate change on human longevity. We find climate change could become the third largest life expectancy reducer behind heart disease and cancer in some European countries. Thus, the cost of inaction on climate change could approach one year of life in some countries."

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
biblio-style: nature
fig_caption: true
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \usepackage[table]{xcolor}
- \usepackage{tabularx}
- \usepackage{makecell}
- \usepackage{xcolor}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[CE,CO]{Climate change life expectancy}
---

* Corresponding author. hauer@uga.edu. p: 706-542-9369.

^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.

^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.

@@

```{r setup, include=FALSE, cache =TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
rm(list=ls())
library(tidyverse)
library(HMDHFDplus)
library(tmap)
library(tmaptools)
library(RColorBrewer)
library(data.table)

###   Inputting user info for HMD

myusername <- "hauer@uga.edu"
mypassword <- "fuckhead"

@@

# Introduction

Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimatin], estimates of displacement [@nicholls2011sea;@hauer2016millions;@hauer2017migration], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate]. As impact quantification studies move further from the physical sciences of climate change into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].

Scholars have long tried to estimate the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer]. Although such estimates are useful, the excess mortality estimates are rather sterile - one death is a tragedy, a million deaths a statistic - and difficult to relate to on an individual level. Life expectancy at birth, on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone. To date, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change. Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating the impact of climate change on human mortality.

Using published data on excess mortality, we connect climate change excess mortality to life expectancy in a mortality model. By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries. This approach allows us to quantify the impact of climate change on human longevity and answer the following question: How much could climate change mortality reduce life expectancy at birth? Our results can situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.

# Methods
We estimate changes in life expectancy by combining three primary datasets: excess mortality data from Forzieri et al. [@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD]. Forzieri et al. [@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B). They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway 2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets. While the disaster data are not standardized across country, Forzeri et al. imputed data in incomplete time periods and countries. This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results. Their results represent business-as-usual climate change and human development without incorporating potential adaptation. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.

To convert excess mortality to life expectancy, we use data gathered from the HMD [@HMD] for corresponding European countries in five-year life tables in conjunction with cause-specific mortality data from the GBD [@GBD;@wang2012age]. The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available. We allocate excess multi-hazard mortality for each country from supplementary table S8 [@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.

We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables (baseline, low, mid, high) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing]. We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$. Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD. Complete life tables for all countries are available in the Supplementary Dataset and all equations are located in the Supplementary Information.

# Results
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig. 1**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.

```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change. We report changes in life expectancies due to climate change for twenty-eight European countries. The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}

@@

```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}

@@

While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig. 2**). The most westerly European countries tend to have the greatest reductions. Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change. Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily]. Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.

# Discussion
In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries. Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality. Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy]. Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer]. Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes. Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].

@@

selectcountries <- c("Spain", "Luxembourg", "Italy", "France", "Portugal", "Ireland", "Slovenia", "Croatia", "Switzerland", "Belgium")
testtbl<- standard %>%
  group_by(COUNTRY) %>%
  gather(COD, ASDR, 2:13) %>%
  filter(!COD %in% c("Circulatorydisease", "Cancer", "ASDR_high", "ASDR_low"),
         !COUNTRY == "Iceland") %>%
  mutate(RANK = rank(desc(ASDR)),
         COD = case_when(
           COD == "Heartdisease" ~ "makecell{Heart\\Disease\\",
           COD == "LungCancer" ~ "makecell{Lung\\Cancer\\",
           COD == "Respiratorydiseases" ~ "makecell{Respiratory\\Diseases\\",
           COD == "Diseasesofthenervoussystem" ~ "makecell{Dis. of the\\Nervous Sys\\",
           COD == "Colorectalcancer" ~ "makecell{Colorectal\\Cancer\\",
           COD == "Suicide" ~ "makecell{Suicide\\",
           COD == "TransportAccidents" ~ "makecell{Transport\\Accidents\\",
           COD == "ASDR_mid"  ~ "cellcolor{blue!25}makecell{cellcolor{blue!25}Climate\\cellcolor{blue!25}Change\\cellcolor{blue!25}"),
         Value = paste0(COD, " ", round(ASDR,2), "}")) %>%
  ungroup()%>%
  arrange(COD, desc(ASDR)) %>%
  #mutate(COUNTRY = paste0(row_number(),COUNTRY))%>%
  select(COUNTRY, Value, RANK) %>%
  spread(RANK, Value) %>%
  slice(match(selectcountries, COUNTRY))

@@

\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} 

@@

Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis. of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\ 
Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis. of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\ 
Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis. of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\ 
France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis. of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\ 
Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis. of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\ 
Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis. of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\ 
Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02} & \makecell{Dis. of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\ 
Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44} & \makecell{Dis. of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\ 
Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis. of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\ 
Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis. of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\ 

>>>>

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "Climate change projected to reduce European life expectancy"
thanks: "All data and code that supports these conclusions are available as supplementary materials."
author:
# - name: Mathew E. Hauer^1^*
#   affiliation: University of Georgia
# - name: Alexis R. Santos^2^
#   affiliation: Pennsylvania State University

abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change [@patz2005impact;@aastrom2013attributing;@forzieri2017increasing;@staddon2014climate]. However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups [@parrish2010peer]. No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy. Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality. In this paper, we convert excess mortality estimates into potential reductions of life expectancy at birth and report the cost of climate change on human longevity. We find climate change could become the third largest life expectancy reducer behind heart disease and cancer in some European countries. Thus, the cost of inaction on climate change could approach one year of life in some countries."

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
biblio-style: nature
fig_caption: true
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \usepackage[table]{xcolor}
- \usepackage{tabularx}
- \usepackage{makecell}
- \usepackage{xcolor}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[C]{Climate change life expectancy}

---

<!-- * Corresponding author. hauer@uga.edu. p: 706-542-9369. -->

<!-- ^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602. -->

<!-- ^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802. -->

@@

```{r setup, include=FALSE, cache =TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
rm(list=ls())
library(tidyverse)
library(HMDHFDplus)
library(tmap)
library(tmaptools)
library(RColorBrewer)
library(data.table)
library(getPass)

###   Inputting user info for HMD
myusername <- getPass(msg = "HMD username: ", noblank = FALSE, forcemask = FALSE)
mypassword <- getPass(msg = "HMD password: ", noblank = FALSE, forcemask = FALSE)

@@

# Main Text

Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimating], estimates of displacement [@rigaud2018groundswell], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate]. As impact quantification studies move further from the physical sciences of climate change into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].

Scholars have long estimated the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer]. Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on an individual level. Life expectancy at birth ($e_0$), on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone. To our knowledge, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change. Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating the impact of climate change on human mortality.

Using published data on excess mortality[@forzieri2017increasing], we connect climate change excess mortality to life expectancy in a mortality model. By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries. This approach allows us to quantify the impact of climate change on human longevity and answer the following questions: What is the cost of inaction on climate change on human longevity? and How does climate change related mortality compare to other mortality vectors? Our results can situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.

<!-- # Methods -->
We estimate changes in life expectancy by combining three primary datasets: excess mortality data due to meterological hazards related to climate change from Forzieri et al. [@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD]. Forzieri et al. [@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B). They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (SSP2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets. While the disaster data are not standardized across country, Forzeri et al. imputed data in incomplete time periods and countries. This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results. Their results represent business-as-usual climate change and human development without incorporating potential adaptation. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.

To convert excess mortality to life expectancy, we use data gathered from the HMD [@HMD] for corresponding European countries in five-year life tables in conjunction with cause-specific mortality data from the GBD [@GBD;@wang2012age]. The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available. We allocate excess multi-hazard mortality for each country from supplementary table S8 in Forzieri et al. [@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.

We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables ($baseline$, $low$, $mid$, $high$) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing]. We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$. Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD. Complete life tables for all countries are available in the Supplementary Dataset and all equations are located in the Supplementary Information.

<!-- # Results -->
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a reduction of up to `r round(lt_lancet$DIF_HIGH[which(lt_lancet$CNTRY == "LUX")],2)*-1` years and the medium-variant of climate hazards could cost the average Spaniard `r round(lt_lancet$DIF_MID[which(lt_lancet$CNTRY == "ESP")],2)*-1` years of life.

```{r figure1, echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change. We report changes in life expectancies due to climate change for twenty-eight European countries. The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.\\label{figure1}")}

@@

```{r figure2, echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.\\label{map}")}

@@

**\autoref{Table1}** reports the rankings of Age Standardised Death Rates (ASDR) for the leading causes of death in these European countries (ASDR's for other CODs come from Eurostat (online data code: hlth_cd_asdr2)) with our estimates of climate change included. In many European countries, climate change could emerge as a top five killer. Only combined CODs, such as all circulatory diseases or all cancers, exhibit higher ASDRs if climatic hazards are more severe than anticipated for Spain, Luxembourg, and France.

While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**\autoref{map}**). The most westerly European countries tend to have the greatest reductions. Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change. Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily]. Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.

<!-- # Discussion -->
In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries. Previous studies on climate change and excess mortality potentially miscommunicate the impact climate change could have on human mortality. Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy]. Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer]. Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes. Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].

@@

selectcountries <- c("Spain", "Luxembourg", "Italy", "France", "Portugal", "Ireland", "Slovenia", "Croatia", "Switzerland", "Belgium")
testtbl<- standard %>%
  group_by(COUNTRY) %>%
  gather(COD, ASDR, 2:13) %>%
  filter(!COD %in% c("Circulatorydisease", "Cancer", "ASDR_high", "ASDR_low"),
         !COUNTRY == "Iceland") %>%
  mutate(RANK = rank(desc(ASDR)),
         ASDR = round(ASDR,2),
         COD = case_when(
           COD == "Heartdisease" ~ "makecell{Heart\\Disease\\",
           COD == "LungCancer" ~ "makecell{Lung\\Cancer\\",
           COD == "Respiratorydiseases" ~ "makecell{Respiratory\\Diseases\\",
           COD == "Diseasesofthenervoussystem" ~ "makecell{Dis. of the\\Nervous Sys\\",
           COD == "Colorectalcancer" ~ "makecell{Colorectal\\Cancer\\",
           COD == "Suicide" ~ "makecell{Suicide\\",
           COD == "TransportAccidents" ~ "makecell{Transport\\Accidents\\",
           COD == "ASDR_mid"  ~ "cellcolor{blue!25}makecell{cellcolor{blue!25}Climate\\cellcolor{blue!25}Change\\cellcolor{blue!25}"),
         Value = paste0(COD, " ", ASDR, "}")) %>%
  ungroup()%>%
  arrange(COD, desc(ASDR)) %>%
  #mutate(COUNTRY = paste0(row_number(),COUNTRY))%>%
  select(COUNTRY, Value, RANK) %>%
  spread(RANK, Value) %>%
  slice(match(selectcountries, COUNTRY))

@@

\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}

@@

Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis. of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\ 
Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis. of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\ 
Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis. of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\ 
France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis. of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\ 
Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis. of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\ 
Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis. of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\ 
Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis. of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\ 
Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 24.44\\\cellcolor{blue!25}\begin{tiny}(13.32 - 38.6)\end{tiny}} & \makecell{Dis. of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\ 
Switzerland & \makecell{Heart\\Disease\\ 97.8} & \makecell{Respiratory\\Diseases\\ 51.3} & \makecell{Dis. of the\\Nervous Sys\\ 44.5} & \makecell{Lung\\Cancer\\ 42.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 22.88\\\cellcolor{blue!25}\begin{tiny}(5.74 - 37.5)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 22.8} & \makecell{Suicide\\ 12.8} & \makecell{Transport\\Accidents\\ 3.6} \\ 
Belgium & \makecell{Respiratory\\Diseases\\ 95.7} & \makecell{Heart\\Disease\\ 72.4} & \makecell{Lung\\Cancer\\ 61.6} & \makecell{Dis. of the\\Nervous Sys\\ 46.5} & \makecell{Colorectal\\Cancer\\ 26.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 21.23\\\cellcolor{blue!25}\begin{tiny}(0.68 - 37.4)\end{tiny}} & \makecell{Suicide\\ 17.3} & \makecell{Transport\\Accidents\\ 6.7} \\ 
